---
title: "Racial Stratification & Utility Burdens — Replication"
author: "Kurt Semm"
date: "`r format(Sys.Date())`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
  echo: true
editor: source
params:
  ipums_xml: "usa_00024.xml"
  ipums_dat: "usa_00024.dat.gz"
  cpi_path:  "CPI.xlsx"
  bins: 34
  burden: "util_burden"    # options: util_burden, water_burden, elec_burden, gas_burden
  keep_other_in_plots: false
  out_dir: "outputs"
  fig_dir: "figs"
---

```{r}
# Allow running chunks interactively: define params if Quarto hasn't
if (!exists("params")) {
  params <- list(
    ipums_xml = "usa_00024.xml",
    ipums_dat = "usa_00024.dat.gz",
    cpi_path  = "CPI.xlsx",
    bins      = 34,
    burden    = "util_burden",
    keep_other_in_plots = FALSE,
    out_dir   = "outputs",
    fig_dir   = "figs"
  )
}

```

# Setup & Reproducibility

```{r}
#| label: setup
# Core packages (keep it lean; avoid fragile theme libs)
library(tidyverse)
library(ipumsr)
library(readxl)
library(data.table)
library(spatstat.geom)   # for weighted histograms (whist)
library(reldist)         # weighted quantiles

set.seed(308)

# Helper: safe factor labeler for IPUMS labelled integers
as_lbl <- function(x) ipumsr::as_factor(x)

# Create output directories if they don't exist
dir.create(params$out_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(params$fig_dir, showWarnings = FALSE, recursive = TRUE)
```

# 1. Data (to match the paper’s pipeline)

## 1.1 Load IPUMS & CPI

```{r}
#| label: load-data
# IPUMS microdata (household-level extracts split by year-block).
# Edit the file paths below to match your downloaded IPUMS files.
# Each row should be one IPUMS extract (XML + DAT.GZ) covering a subset of years.
extracts <- tibble::tibble(
  xml = c(
    "usa_00026.xml",
    "usa_00027.xml",
    "usa_00028.xml"
  ),
  dat = c(
    "usa_00026.dat.gz",
    "usa_00027.dat.gz",
    "usa_00028.dat.gz"
  )
)

# Read and stack the microdata from each extract, keeping only the variables
# actually used later in the script to reduce memory usage.
cps_list <- purrr::map2(
  extracts$xml,
  extracts$dat,
  ~ {
    ddi <- read_ipums_ddi(.x)
    read_ipums_micro(ddi, data_file = .y, verbose = FALSE) %>%
      dplyr::select(
        YEAR, HHWT,
        HHINCOME, COSTWATR, COSTELEC, COSTGAS,
        FARM, PLUMBING, SINK, HOTWATER, FOODSTMP,
        RELATE, COUNTYFIP,
        PERNUM, EDUC, AGE, RACE, HISPAN, SEX
      )
  }
)

cps_data <- dplyr::bind_rows(cps_list)
rm(cps_list)
gc()

# CPI with your actual column names
CPI <- read_excel(params$cpi_path)

cpi <- CPI %>%
  rename(
    YEAR   = Year,
    Annual = `Annual Average CPI`
  ) %>%
  mutate(
    YEAR = as.integer(YEAR)
  ) %>%
  select(YEAR, Annual)

```

## 1.2 Cleaning & Sample Restrictions

**Target unit**: head of household; civilian/non-farm; complete plumbing and sink; valid county; positive income and water cost.

```{r}
#| label: clean-sample

cps <- cps_data %>%
  # 1) Join CPI and drop years without CPI
  left_join(cpi, by = "YEAR") %>%
  filter(!is.na(Annual)) %>%

  # 2) Attach factor labels for filters
  mutate(
    PLUMBING_f = as_lbl(PLUMBING),
    SINK_f     = as_lbl(SINK),
    HOTWATER_f = as_lbl(HOTWATER),
    FOODSTMP_f = as_lbl(FOODSTMP)
  ) %>%

  # 3) Apply your sample restrictions
  filter(
    COSTWATR >= 1,
    HHINCOME >= 1,
    FARM == 1,                    # 1 = non-farm household (IPUMS coding)
    PLUMBING_f == "Complete plumbing",
    SINK_f == "Has sink",
    RELATE == 1,                  # head of household
    !is.na(COUNTYFIP)
  ) %>%

  # 4) Deflate to constant dollars (base = 100)
  mutate(
    HHINCOME = (HHINCOME/Annual)*100,
    COSTWATR = (COSTWATR/Annual)*100,
    COSTELEC = (COSTELEC/Annual)*100,
    COSTGAS  = (COSTGAS /Annual)*100
  )
```

## 1.3 Household-size Equivalization

We equivalize **income and utility costs** by household size with exponent \\(n=0.5\\).

```{r}
#| label: equivalize
n_eq <- 0.5
cps <- cps %>%
  mutate(
    HHINCOME = HHINCOME/(PERNUM^n_eq),
    COSTWATR = COSTWATR/(PERNUM^n_eq),
    COSTELEC = COSTELEC/(PERNUM^n_eq),
    COSTGAS  = COSTGAS /(PERNUM^n_eq)
  )
```

## 1.4 Construct Utility Burden Outcomes

```{r}
#| label: construct-burdens
cps <- cps %>%
  mutate(
    util_burden  = (COSTWATR + COSTELEC + COSTGAS)/HHINCOME,
    water_burden = COSTWATR/HHINCOME,
    elec_burden  = COSTELEC/HHINCOME,
    gas_burden   = COSTGAS/HHINCOME
  )
```

## 1.5 Identity & Demographics (Education, Age, Race×Sex)

```{r}
#| label: identity-groups
cps <- cps %>%
  mutate(
    EDUC_f = as_lbl(EDUC),
    EDUC4 = case_when(
      EDUC_f %in% c("N/A or no schooling","Nursery school to grade 4",
                     "Grade 5, 6, 7, or 8","Grade 9","Grade 10","Grade 11") ~ "<HS",
      EDUC_f == "Grade 12" ~ "HS",
      EDUC_f %in% c("1 year of college","2 years of college","3 years of college") ~ "SomeCollege",
      EDUC_f %in% c("4 years of college","5+ of college") ~ "CollegePlus",
      TRUE ~ NA_character_
    ),
    AGE = as.numeric(AGE),
    AGE_GRP = cut(AGE, breaks = c(18,29,39,49,59,64), include.lowest = TRUE,
                  labels = c("20","30","40","50","60")),
    RACE_f = as_lbl(RACE),
    HISPAN_f = as_lbl(ipumsr::lbl_na_if(HISPAN, ~ .val == 9)),
    SEX_f = as_lbl(SEX),
    RACE_ETH = case_when(
      HISPAN_f %in% c("Mexican","Puerto Rican","Cuban","Other") ~ "Hispanic",
      HISPAN_f == "Not Hispanic" & RACE_f == "White" ~ "White",
      HISPAN_f == "Not Hispanic" & RACE_f == "Black/African American" ~ "Black",
      HISPAN_f == "Not Hispanic" & RACE_f == "American Indian or Alaska Native" ~ "American Indian",
      TRUE ~ "Other"
    ),
    IDENTITY = paste(RACE_ETH, SEX_f)
  ) %>%
  filter(!is.na(EDUC4), !is.na(IDENTITY))
```

# 2. Binning & Weighted Histograms

Per **YEAR**, drop the top 2% by outcome, then create **34 equal-width bins** from 0 to the year’s 98th percentile cap. Use **HHWT** for weights.

```{r}
#| label: caps-and-bins

# 1. Weighted 98th percentiles for each outcome
wq <- function(x, q, w) {
  # handle all-NA quietly
  if (all(is.na(x))) return(NA_real_)
  reldist::wtd.quantile(x, q = q, weight = w)
}

caps <- cps %>%
  group_by(YEAR) %>%
  summarise(
    util_98  = wq(util_burden,  0.98, HHWT),
    water_98 = wq(water_burden, 0.98, HHWT),
    elec_98  = wq(elec_burden,  0.98, HHWT),
    gas_98   = wq(gas_burden,   0.98, HHWT),
    .groups = "drop"
  )

cps <- cps %>% left_join(caps, by = "YEAR")

# 2. Robust histogram helper
make_hist <- function(x, top, w, k = params$bins) {
  # top may come in as length 0 or >1; squash to first element
  top <- top[1]

  # guard against bad caps (NA, non-finite, non-positive, length 0)
  if (length(top) == 0 || is.na(top) || !is.finite(top) || top <= 0) {
    return(tibble(
      bin  = factor(),
      mid  = numeric(),
      mass = numeric(),
      p    = numeric()
    ))
  }

  brks <- seq(0, top, length.out = k + 1)
  freq <- as.numeric(spatstat.geom::whist(x, breaks = brks, weights = w))

  tibble(
    bin   = factor(seq_len(k)),
    mid   = brks[-1] - diff(brks)/2,
    mass  = freq,
    p     = if (sum(freq) > 0) freq / sum(freq) else freq
  )
}

# 3. Choose burden and matching cap
burden_var <- rlang::sym(params$burden)

cap_name <- paste0(
  stringr::str_remove(params$burden, "_burden"),
  "_98"
)
cap_var <- rlang::sym(cap_name)

# 4. Conditional distributions f(Y | EDUC4, IDENTITY)
hist_df <- cps %>%
  # drop rows with missing burden or missing cap
  filter(!is.na(!!burden_var), !is.na(!!cap_var)) %>%
  group_by(YEAR, EDUC4, IDENTITY) %>%
  group_modify(~ make_hist(
    x   = .x[[rlang::as_string(burden_var)]],
    top = .x[[rlang::as_string(cap_var)]],
    w   = .x$HHWT
  )) %>%
  ungroup()

# 5. Unconditional f(Y) per year
fy <- cps %>%
  filter(!is.na(!!burden_var), !is.na(!!cap_var)) %>%
  group_by(YEAR) %>%
  group_modify(~ make_hist(
    x   = .x[[rlang::as_string(burden_var)]],
    top = .x[[rlang::as_string(cap_var)]],
    w   = .x$HHWT
  )) %>%
  ungroup() %>%
  select(YEAR, bin, p)

# (optional) Save histogram data
readr::write_csv(hist_df, file.path(params$out_dir, paste0("hist_", params$burden, ".csv")))
readr::write_csv(fy,      file.path(params$out_dir, paste0("fy_",   params$burden, ".csv")))
```

# 3. Information Measures

We use natural logs. Define entropy \\(H\\), unconditional \\(H(Y)\\), and conditional \\(H(Y\|G)\\). Normalized associations use division by \\(H(Y)\\).

```{r}
#| label: info-measures
entropy <- function(p){ p <- p[p > 0 & is.finite(p)]; -sum(p * log(p)) }

HY <- fy %>% group_by(YEAR) %>% summarise(HY = entropy(p), .groups="drop")
readr::write_csv(HY, file.path(params$out_dir, paste0("HY_", params$burden, ".csv")))

# H(Y|EDUC4)
HY_given_educ <- hist_df %>%
  group_by(YEAR, EDUC4) %>%
  summarise(H = entropy(p), .groups="drop") %>%
  left_join(HY, by = "YEAR") %>%
  mutate(a_y_educ = (HY - H)/HY)
readr::write_csv(HY_given_educ, file.path(params$out_dir, paste0("HY_given_educ_", params$burden, ".csv")))

# H(Y|IDENTITY)
HY_given_ident <- hist_df %>%
  group_by(YEAR, IDENTITY) %>%
  summarise(H = entropy(p), .groups="drop") %>%
  left_join(HY, by = "YEAR") %>%
  mutate(a_y_ident = (HY - H)/HY)
readr::write_csv(HY_given_ident, file.path(params$out_dir, paste0("HY_given_ident_", params$burden, ".csv")))

# H(Y|EDUC4, IDENTITY) and a(Y||(EDUC,IDENT))
HY_given_ei <- hist_df %>%
  group_by(YEAR, EDUC4, IDENTITY) %>%
  summarise(H = entropy(p), .groups="drop") %>%
  left_join(HY, by = "YEAR") %>%
  mutate(a_y_ei = (HY - H)/HY)
readr::write_csv(HY_given_ei, file.path(params$out_dir, paste0("HY_given_ei_", params$burden, ".csv")))

# Tripartite interaction m(Y, EDUC, IDENTITY) = a(Y||EDUC) + a(Y||IDENT) - a(Y||(EDUC,IDENT))
m_tripartite <- HY_given_ei %>%
  left_join(HY_given_educ %>% select(YEAR, EDUC4, a_y_educ), by = c("YEAR","EDUC4")) %>%
  left_join(HY_given_ident %>% select(YEAR, IDENTITY, a_y_ident), by = c("YEAR","IDENTITY")) %>%
  transmute(YEAR, EDUC4, IDENTITY, m = a_y_educ + a_y_ident - a_y_ei)
readr::write_csv(m_tripartite, file.path(params$out_dir, paste0("m_tripartite_", params$burden, ".csv")))
```

# 4. Plots (mirroring Wiener & dos Santos, pp. 26–39)

## 4.1 a(Y \|\| Education) over time

```{r}
#| label: plot-educ
# Optionally reorder panels
educ_levels <- c("<HS","HS","SomeCollege","CollegePlus")
HY_given_educ <- HY_given_educ %>% mutate(EDUC4 = factor(EDUC4, levels = educ_levels))

p_educ <- ggplot(HY_given_educ, aes(YEAR, a_y_educ, group = EDUC4)) +
  geom_line() + geom_point() +
  facet_wrap(~EDUC4, nrow = 1) +
  labs(x = NULL, y = "a(Y || Education)",
       title = paste0("Informational association of ", params$burden, " with Education")) +
  theme_minimal()

print(p_educ)
ggsave(filename = file.path(params$fig_dir, paste0("educ_", params$burden, ".png")),
       plot = p_educ, width = 10, height = 4, dpi = 300)
```

## 4.2 a(Y \|\| Identity) over time

```{r}
#| label: plot-ident
id_df <- if (isTRUE(params$keep_other_in_plots)) HY_given_ident else HY_given_ident %>% filter(!grepl("^Other", IDENTITY))

p_ident <- ggplot(id_df, aes(YEAR, a_y_ident, color = IDENTITY, group = IDENTITY)) +
  geom_line() + geom_point() +
  labs(x = NULL, y = "a(Y || Identity)",
       title = paste0("Informational association of ", params$burden, " with Race×Sex")) +
  theme_minimal()

print(p_ident)
ggsave(filename = file.path(params$fig_dir, paste0("identity_", params$burden, ".png")),
       plot = p_ident, width = 10, height = 5, dpi = 300)
```

## 4.3 Tripartite m(Y, EDUC, IDENTITY)

```{r}
#| label: plot-m
m_df <- if (isTRUE(params$keep_other_in_plots)) m_tripartite else m_tripartite %>% filter(!grepl("^Other", IDENTITY))

p_m <- ggplot(m_df, aes(YEAR, m, color = IDENTITY)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_line() + geom_point() +
  facet_wrap(~EDUC4) +
  labs(x = NULL, y = "m(Y, Educ, Identity) — synergy if m < 0",
       title = paste0("Interaction information for ", params$burden)) +
  theme_minimal()

print(p_m)
ggsave(filename = file.path(params$fig_dir, paste0("m_tripartite_", params$burden, ".png")),
       plot = p_m, width = 10, height = 6, dpi = 300)
```

# 5. (Optional) Component Analyses

Set `params$burden` to `"water_burden"`, `"elec_burden"`, or `"gas_burden"` and re-knit to generate the same figure set by component.

# Appendix: Session Info

```{r}
sessionInfo()
```